Model Finding for Recursive Functions in SMT

### 2016-01-26 ijcar2016/body.tex
break
Fix results rows
some edits (mostly on 4.1)

\title{Model Finding for Recursive Functions in SMT\thanks{%
This research is partially supported by the Inria technological development
action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC).
}
}
\titlerunning {Model Finding for Recursive Functions in SMT}

@@

a model of formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $n$ must also satisfy~$\psi$:

@@

\[\con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \smash{\bigwedge\nolimits_{\vthinspace i=0}^{\!n-1}} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq \con{c}-i + \con{s}( \con{c}-i-1 ) \bigr)
\wedge \con{c}-n \leq 0 \wedge \con{s}( \con{c}-n ) \teq 0\]

@@

We have implemented the translation $\conv$ in two separate systems, 
as a preprocessor in the development version of \cvc (version~1.5 prerelease),
and in the higher-order model finder Nunchaku.
This section describes how the translation is implemented in both these systems,
as well as optimizations used by \cvc for finding models for the translated benchmarks.

@@

TODO:
When invoked with a specific
command-line option, the solver assumes that functions introduced using the
\texttt{define-fun-rec} command are admissible and translates them accordingly.

In \cvc, function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written using the $\definefunreccmd$ command
from the SMT-LIB 2.5 standard~\cite{smtlib25}.
Formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as %follows:
%
\begin{quote}
%\small -- no small, unless you convince me that these mixtures of font sizes
% are aesthetic and serve a purpose --jb
\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (+ x (s (- x 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\begin{rep}%
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
\end{rep}%
\begin{conf}%
\cvc adds
$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$
to its list of assertions,
which after rewriting becomes
$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$. \end{conf}%
If \cvc's finite model finding mode for recursive functions is enabled (using
the command-line option \texttt{-}\texttt{-fmf-fun}), it will replace its list of known
assertions based on the translation $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. 
For the example above, it outputs a %partial
model of~$\con{s}$ where only
the values of $\con{s}(x)$ for $x = 0,\ldots,14$ are correctly given:
\begin{quote}
%\small
\begin{verbatim}
(model
  (define-fun s (($x1 Int)) Int 
    (ite (= $x1 14) 105 (ite (= $x1 13) 91 (ite (= $x1 12) 78 
      (ite (= $x1 11) 66 (ite (= $x1 10) 55 (ite (= $x1 4) 10 
        (ite (= $x1 9) 45 (ite (= $x1 8) 36 (ite (= $x1 7) 28 
          (ite (= $x1 6) 21 (ite (= $x1 3) 6 (ite (= $x1 5) 15 
            (ite (= $x1 2) 3 (ite (= $x1 1) 1 0)))))))))))))))
  (define-fun c () Int 14))
\end{verbatim}
\end{quote}

The \texttt{-}\texttt{-fmf-fun} option tells \cvc to assume that
functions introduced using \texttt{define-\allowbreak fun-\allowbreak rec} are admissible.
We stress that admissibility must be discharged separately by the user---e.g., using a
syntactic criterion or a termination prover.
If some function definitions are not admissible, \cvc may answer \emph{sat} for
an unsatisfiable problem. 
\begin{rep}
Indeed, if we add the inconsistent definition
%
\begin{quote}
%\small
\begin{verbatim}
(define-fun-rec h ((x Int)) Int (+ (h x) x))
\end{verbatim}
\end{quote}
%
to the above problem and run \cvc with the \texttt{-}\texttt{-fmf-fun} option,
it wrongly answers \emph{sat}.
\end{rep}

\paragraph{Optimizations for Finding Finite Models of $\conv(\varphi)$}
Like other systems,
the finite model finding capability of \cvc incrementally fixes bounds on the cardinalities of uninterpreted types,
and increases these bounds until it encounters a model.
When multiple sorts are present, it uses a fairness scheme which
bounds the sum of cardinalities of all uninterpreted types~\cite{reynolds2013finite}.
For instance, if a signature has two uninterpreted types $\tau_1$ and $\tau_2$,
it will first search for models where 
the cardinality of $\tau_1$ plus the cardinality of $\tau_2$ is at most $2$,
then $3$, $4$, and so on.
To accelerate the search for models,
we use an optimization based on statically inferring \emph{monotone} types.
A monotone type is one in which models can always be extended with an additional element of that type.
Inference of monotone types has been show in previous work~\cite{claessen2011mono,korovin-2013}
to be an important component of finite model finding systems.
Moreover, it is easy to see that types $\fargtype{\con{f}}$ introduced by our translation $\conv$ are monotone.
The finite model finding capability of \cvc may take advantage of this fact by fixing the bounds for all 
monotone types simultaneously.
That is, if $\tau_1$ and $\tau_2$ are both monotone types (either introduced by our translation, or are otherwise inferred to be),
then we fix the bound for \emph{both} of these types to be $1$, then $2$, and so on.
This scheme allows the solver greater flexibility with respect to the default scheme,
and comes with no loss of generality with respect to models, since monotone sorts can always be extended to have equal cardinalities.

@@

\begin{figure}[t]
\small
%\normalsize
%\small
\centering
\begin{tabular}{l@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\,\,}}
  & \multicolumn{2}{c@{\kern1.5em}}{\phantom{0}\ziiib}     & \multicolumn{2}{c@{\kern1.5em}}{\phantom{0}\cvcd}     
  & \multicolumn{2}{c@{\kern1.5em}}{\phantom{0}\cvcf}      & \multicolumn{2}{c@{\kern1.5em}}{\cvcfe} 
  & \multicolumn{2}{c@{\,\,}}{\cvcfm}
\\%[-1pt]
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
\\
\midrule
\isanun & {\win 41} & 40 & 28 & 34 & 26 & 39 & 26 & 39 & 26 & 39
\\
\isa & 14 & {\win 15} & {\win 15} & {\win 15} & 1 & {\win 15} & {\win 15} & {\win 15} & 1 & {\win 15}
\\
\leon & 54 & 55 & {\win 56} & {\win 56} & 13 & 55 & {\win 56} & 55 & 13 & 55
\\
\isam & 48 & 52 & {\win 58} & 52 & 12 & 53 & {\win 58} & 50 & 12 & 53
\\
\leonm & 94 & 107 & 106 & 107 & 39 & {\win 108} & 106 & {\win 108} & 39 & {\win 108}
\\[\jot]
Total & 251 & 269 & 263 & 264 & 91 & {\win 270} & 261 & 267 & 91 & {\win 270}
\end{tabular}
\caption{\,Number of \emph{unsat} responses on benchmarks without and with $\conv$ translation}
\label{fig:unsat}
\end{figure}

@@

{%\footnotesize
\def\ackname{Acknowledgment}
\paragraph{%\footnotesize
\ackname.}
Viktor Kuncak and Stephan Merz have made this work possible. We would also like
to thank Damien Busato-Gaston and Emmanouil Koukoutos for providing the
set of Leon benchmarks used in the evaluation, and the anonymous reviewers for their
suggestions and comments.
}
%

>>>>

\title{Model Finding for Recursive Functions in SMT}
\titlerunning {Model Finding for Recursive Functions in SMT}

@@

a model of formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as~$n$ must also satisfy~$\psi$:

@@

\[\begin{aligned}[t]
\con{s}( \con{c} ) > 100 %\displaystyle
& {} \wedge \relax{\bigwedge\nolimits_{\vthinspace i=0}^{\!\smash{n-1}}} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq \con{c}-i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
& {} \wedge \con{c}-n \leq 0 \wedge \con{s}( \con{c}-n ) \teq 0
\end{aligned}\]

@@

We have implemented the translation $\conv$ in two separate systems, 
as a preprocessor in \cvc (version~1.5 prerelease)
and in the \cvc-based higher-order model finder Nunchaku.
This section describes how the translation is implemented in these systems,
as well as optimizations used by \cvc for finding models of translated
problems. % for the translated benchmarks.

@@

% When invoked with a specific
% command-line option, the solver assumes that functions introduced using the
% \texttt{define-fun-rec} command are admissible and translates them accordingly.

In \cvc, function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written using the $\definefunreccmd$ command
from SMT-LIB 2.5~\cite{smtlib25}.
Formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as %follows:
%
\begin{quote}
%\small -- no small, unless you convince me that these mixtures of font sizes
% are aesthetic and serve a purpose --jb
\verb|(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (+ x (s (- x 1)))))|\kern-200mm \\
\verb|(declare-fun c () Int)| \\
\verb|(assert (> (s c) 100))| \\
\verb|(check-sat)|
\end{quote}
%
When reading this input,
\begin{rep}%
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
\end{rep}%
\begin{conf}%
\cvc adds
$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$
to its list of assertions,
which after rewriting becomes
$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$. \end{conf}%
By specifying the command-line option \texttt{-}\texttt{-fmf-fun}, users
can enable \cvc's finite model finding mode for recursive functions.
In this mode, \cvc will replace its list of known
assertions based on the $\conv$ translation before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. 
For the example above, it outputs a %partial
model of~$\con{s}$ where only
the values of $\con{s}(x)$ for $x = 0,\ldots,14$ are correctly given:
\begin{quote}
%\small
\begin{verbatim}
(model
  (define-fun s (($x1 Int)) Int 
    (ite (= $x1 14) 105 (ite (= $x1 13) 91 (ite (= $x1 12) 78 
      (ite (= $x1 11) 66 (ite (= $x1 10) 55 (ite (= $x1 4) 10 
        (ite (= $x1 9) 45 (ite (= $x1 8) 36 (ite (= $x1 7) 28 
          (ite (= $x1 6) 21 (ite (= $x1 3) 6 (ite (= $x1 5) 15 
            (ite (= $x1 2) 3 (ite (= $x1 1) 1 0)))))))))))))))
  (define-fun c () Int 14))
\end{verbatim}
\end{quote}

With the \texttt{-}\texttt{-fmf-fun} option enabled, \cvc assumes that
functions introduced using \texttt{define-\allowbreak fun-\allowbreak rec} are admissible.
Admissibility must be discharged separately by the user---e.g., using a
syntactic criterion or a termination prover.
If some function definitions are not admissible, \cvc may answer \emph{sat} for
an unsatisfiable problem. 
\begin{rep}
Indeed, if we add the inconsistent definition
%
\begin{quote}
%\small
\begin{verbatim}
(define-fun-rec h ((x Int)) Int (+ (h x) x))
\end{verbatim}
\end{quote}
%
to the above problem and run \cvc with the \texttt{-}\texttt{-fmf-fun} option,
it wrongly answers \emph{sat}.
\end{rep}

\cvc implements a few optimizations designed to help finding finite models of
$\conv(\varphi)$.
Like other systems,
the finite model finding capability of \cvc incrementally fixes bounds on the cardinalities of uninterpreted types,
and increases these bounds until it encounters a model.
When multiple sorts are present, it uses a fairness scheme which
bounds the sum of cardinalities of all uninterpreted types~\cite{reynolds2013finite}.
For instance, if a signature has two uninterpreted types $\tau_1$ and $\tau_2$,
it will first search for models where 
the cardinality of $\tau_1$ plus the cardinality of $\tau_2$ is at most $2$,
then $3$, $4$, and so on.
To accelerate the search for models,
we use an optimization based on statically inferring \emph{monotone} types.
A monotone type is one in which models can always be extended with an additional element of that type.
Inference of monotone types has been show in previous work~\cite{claessen2011mono,korovin-2013}
to be an important component of finite model finding systems.
Moreover, it is easy to see that types $\fargtype{\con{f}}$ introduced by our translation $\conv$ are monotone.
The finite model finding capability of \cvc may take advantage of this fact by fixing the bounds for all 
monotone types simultaneously.
That is, if $\tau_1$ and $\tau_2$ are both monotone types (either introduced by our translation, or are otherwise inferred to be),
then we fix the bound for \emph{both} of these types to be $1$, then $2$, and so on.
This scheme allows the solver greater flexibility with respect to the default scheme,
and comes with no loss of generality with respect to models, since monotone sorts can always be extended to have equal cardinalities.

@@

\begin{figure}[t]
\small
%\normalsize
%\small
\centering
\begin{tabular}{l@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\kern1.5em}r@{\kern0.375em}r@{\,\,}}
  & \multicolumn{2}{c@{\kern1.5em}}{\phantom{0}\ziiib}     & \multicolumn{2}{c@{\kern1.5em}}{\phantom{0}\cvcd}     
  & \multicolumn{2}{c@{\kern1.5em}}{\phantom{0}\cvcf}      & \multicolumn{2}{c@{\kern1.5em}}{\cvcfe} 
  & \multicolumn{2}{c@{\,\,}}{\cvcfm}
\\%[-1pt]
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
  & \hfill $\varphi$ \hfill & $\conv(\varphi)$\!\!
\\
\midrule
\isanun & {\win 41} & 40 & 28 & 34 & 26 & 39 & 26 & 39 & 26 & 39
\\
\isa & 14 & {\win 15} & {\win 15} & {\win 15} & 1 & {\win 15} & {\win 15} & {\win 15} & 1 & {\win 15}
\\
\isam & 54 & 55 & {\win 56} & {\win 56} & 13 & 55 & {\win 56} & 55 & 13 & 55
\\
\leon & 48 & 52 & {\win 58} & 52 & 12 & 53 & {\win 58} & 50 & 12 & 53
\\
\leonm & 94 & 107 & 106 & 107 & 39 & {\win 108} & 106 & {\win 108} & 39 & {\win 108}
\\[\jot]
Total & 251 & 269 & 263 & 264 & 91 & {\win 270} & 261 & 267 & 91 & {\win 270}
\end{tabular}
\caption{\,Number of \emph{unsat} responses on benchmarks without and with $\conv$ translation}
\label{fig:unsat}
\end{figure}

@@

{%\footnotesize
\def\ackname{Acknowledgment}
\paragraph{%\footnotesize
\ackname.}
Viktor Kuncak and Stephan Merz have made this work possible. We would also like
to thank Damien Busato-Gaston and Emmanouil Koukoutos for providing the
set of Leon benchmarks used in the evaluation, and the anonymous reviewers for their
suggestions and comments. Cruanes is supported by the Inria technological
development action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC).
}
%
